Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-10-24 draft.tex
.

We used a leaky integrate-and-fire-model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
dV = -\frac{V-E_l}{\tau_m}dt + \frac{\sigma}{\sqrt{\tau_m}}dW
\label{LIF_Dynamics}
\end{equation}
where $V$ is the membrane potential, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron by $w^{effective}_{ji}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}
\begin{tabular}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabular}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first attempt, we set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after 700 seconds, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group, the average $NO$ concentration at the excitatory neurons' positions and the average firing threshold within the exc. group (inh. threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in the upper three plots of Figure \ref{full_sim_osci}, are in fact regular oscillations across all three depicted variables. While the oscillation amplitudes undergo a rather unpredictable time course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. Although one might argue that - for the sake of practical purposes - regular oscillations of the $NO$ concentration are not of much concern for the network as such, a periodically fluctuating firing rate is a qualitative feature requiring further inquiry.

\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at exc. neurons' positions. C/C*: Average exc. firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}

@@

where $\tau_e$ and $\tau_i$ represents time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. According to the Poincaré-Bendixson Theorem, such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing "into" $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.
Coming back to the dynamics in question, one would need to show that the network has a locally unstable fixed point at $\mathrm{(r_e,r_i)\approx (3\, Hz, 13\, Hz)}$, see A* in Figure \ref{full_sim_osci}. Several aspects come into play, making it a hard task to fit the full spiking network model onto the simplified equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. First, the choice of $\tau_e$ and $\tau_i$ in \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} has significant influence onto the stability of the system. However, a straightforward equivalent does not exist in the used spiking neuron model. It has been shown that in the case of slowly varying input $\tau_{e/i}$ of equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} are equal to the membrane constant $\tau_m$ of the LIF-neuron used in the network \cite{Gerstner_Pop_Act}, see equation \ref{LIF_Dynamics}. A second problem is to choose a good estimate of the activation functions $\phi_{e/i}( \cdot )$. In the absence of noise in equation \ref{LIF_Dynamics}, the firing rate for a given constant total input $J$ is
\begin{equation}
\phi(J) = \begin{cases}\frac{1}{-\tau_m ln(\frac{V_t -E_l-J\tau_m}{V_r -E_l-J\tau_m})} & \mathrm{for}\; J\tau_m + E_l > V_t\\
0 & \mathrm{otherwise}
\end{cases}.
\label{LIF_Fir_Rate_no_Noise}
\end{equation}
If membrane noise is present, according to \cite{Roxin_Firing_Rate_Distribution} one can calculate the \emph{mean} firing rate by
\begin{equation}
\phi(J) = 
\label{LIF_Fir_Rate_with_Noise}
\end{equation}

>>>>

We used a leaky integrate-and-fire-model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV = -(V-E_l)dt + \sqrt{\tau_m} \sigma dW
\label{LIF_Dynamics}
\end{equation}
where $V$ is the membrane potential, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron by $w^{effective}_{ji}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}
\begin{tabular}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabular}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first attempt, we set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after 750 seconds, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group, the average $NO$ concentration at the excitatory neurons' positions and the average firing threshold within the exc. group (inh. threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in the upper three plots of Figure \ref{full_sim_osci}, are in fact regular oscillations across all three depicted variables. While the oscillation amplitudes undergo a rather unpredictable time course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. Although one might argue that - for the sake of practical purposes - regular oscillations of the $NO$ concentration are not of much concern for the network as such, a periodically fluctuating firing rate is a qualitative feature requiring further enquiry.
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at exc. neurons' positions. C/C*: Average exc. firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-750\, s}$ , diffusive h. for $\mathrm{750-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}

@@

where $\tau_e$ and $\tau_i$ represents time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. According to the Poincaré-Bendixson Theorem, such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing "into" $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.
Coming back to the dynamics in question, one would need to show that the network has a locally unstable fixed point at $\mathrm{(r_e,r_i)\approx (3\, Hz, 13\, Hz)}$, see A* in Figure \ref{full_sim_osci}. Several aspects come into play, making it a hard task to fit the full spiking network model onto the simplified equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. First, the choice of $\tau_e$ and $\tau_i$ in \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} has significant influence onto the stability of the system. However, a straightforward equivalent does not exist in the used spiking neuron model. It has been shown that in the case of slowly varying input $\tau_{e/i}$ of equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} are equal to the membrane constant $\tau_m$ of the LIF-neuron used in the network \cite{Gerstner_Pop_Act}, see equation \ref{LIF_Dynamics}. A second problem is to choose a good estimate of the activation functions $\phi_{e/i}( \cdot )$. In the absence of noise in equation \ref{LIF_Dynamics}, the firing rate for a given constant total input $J$ is
\begin{equation}
\phi(J) = \begin{cases}\frac{1}{-\tau_m ln(\frac{V_t -E_l-J\tau_m}{V_r -E_l-J\tau_m})} & \mathrm{for}\; J\tau_m + E_l > V_t\\
0 & \mathrm{otherwise}
\end{cases}.
\label{LIF_Fir_Rate_no_Noise}
\end{equation}
If noise $\sigma$ is present, according to \cite{Roxin_Firing_Rate_Distribution} one can calculate the \emph{mean} firing rate by
\begin{align}
\phi(J) &= \left[ \sqrt{\pi}\tau_m \int_{x_-}^{x_+} dx e^{x^2} erfc(-x) \right]^{-1} \label{LIF_Fir_Rate_with_Noise1}\\
x_- &= (V_r-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise2}\\
x_+ &= (V_t-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise3}
\end{align}
where $erfc$ is the complementary error function. Figure \ref{F-I_noise_dep} shows that both predictions are quite accurate in predicting the mean firing rate. 
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/F-I_noise_dep.png}
\end{center}
\caption{Comparison of mean firing rate of a LIF neuron without noise (squares) and with noise (circles, $\mathrm{\sigma = \sqrt{5}\, mV}$), averaged over $\mathrm{50\, s}$. Red and blue curves are the corresponding analytic predictions, see equation \ref{LIF_Fir_Rate_no_Noise} and \ref{LIF_Fir_Rate_with_Noise1}.}
\label{F-I_noise_dep}
\end{figure}
However, one should note that the synaptic input within a spiking network carries an intrinsic randomness: since it is the sum of instantaneous increases in membrane voltage upon arriving spikes, one cannot simply describe it as a mean constant input $J$. Rather, if we assume that the arrival of spikes is approximately poisson-distributed and the time constant of the membrane is small compared to the average interspike-interval, according to \cite{Roxin_Firing_Rate_Distribution}, one can describe the synaptic input from population $\mathrm{A}$ to population $\mathrm{B}$ as $J_{AB}(t) = \mu_{AB} (t) + \sigma_{AB} \cdot \zeta(t)$, where $\zeta(t)$ is zero-mean standard gaussian noise and
\begin{align}
\mu_{AB} &= <w_{AB}> \cdot <N_{in,AB}> \cdot <r_A> \label{Noise_Approx_1}\\
\sigma_{AB} &= \sqrt{\tau_m \cdot <w_{AB}>^2 \cdot <N_{in,AB}> \cdot <r_A>} \label{Noise_Approx_1} \, .
\end{align}
$\mathrm{<w_{AB}>,\, <N_{in,AB}>}$ and $\mathrm{<r_A>}$ denote the mean weight of a synapse connection population $\mathrm{A}$ with population $\mathrm{B}$, the mean number of incoming connections at a neuron in population $\mathrm{B}$ and the mean firing rate in population $\mathrm{A}$ (which makes it necessary to assume changes within the firing rate to be slow enough for a quasi-stationary description of the activity).
Since we can assume intrinsic and input noise to be uncorrelated, the total noise within a neuron of population $\mathrm{B}$ has a standard deviation of 
\begin{equation}
\sigma_{tot.} = \sqrt{\sigma_{intr.}^2 + \sum_A \sigma_{AB}^2} \; .
\label{Total_Noise_Neuron}
\end{equation}

Apart from the correct description of $\phi_{e/i}$, a third problem is the analytic description of the STP that was present for recurrent excitatory connections in our network. In \cite{Markram_STP}, the steady-state value of $\mathrm{x\cdot u}$ for an input of constant rate $\mathrm{r}$ is given by
\begin{align}
x_{st.}(r) &= \frac{1-\exp \left(-\frac{1}{r\cdot \tau _d}\right)}{1-\left(1-u_{st.}\left(r\right)\right)\cdot \exp \left(-\frac{1}{r\cdot \tau _d}\right)} \label{STP_steady1} \\
u_{st.}(r) &= \frac{U}{1-\left(1-U\right)\exp \left(-\frac{1}{r\cdot \tau _f}\right)} \; .\label{STP_steady2}
\end{align}
When trying to find the fixed point of equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, one thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations results in a stable configuration depicted in \ref{Dyn_Rate_Approx}. For this simulation, we set the relevant parameters to their predefined values and, in addition, extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate between 100-750 s of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that where present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principal, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: Assuming the validity of the described rate-model, the recurrent network as such (without homeostasis) is stable under the given choice of parameters.
